---
title: "Lab 02 - Global plastic waste"
author: "KEY"
date: "`r Sys.Date()`"
output: html_document
---

### Load packages and data

```{r load-packages, message = FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```


## Load & join data sets

```{r load-data, message = FALSE}
# Member 2 (pair 1) write code to load File 1 data here:
data1 <- read_csv("data/mismanaged-plastic-waste-per-capita.csv")

# Member 4 (pair 2) write code to load File 2 data here:
data2 <- read_csv2("data/per-capita-ocean-plastic-waste.txt")

# Member 1 (pair 1) write code to load File 3 data here:
data3 <- read_tsv("data/UN_country_population.tsv")

# Member 3 (pair 2) write code to load File 4 data here:
data4 <- read_excel("data/UN_country_region.xlsx")

```



```{r join-data, message = FALSE}
# Member 2 (pair 1) write code to create object 'plastic_data_all':
plastic_data_all <- data1

# Member 4 (pair 2) write code to join your data to 'plastic_data_all':
plastic_data_all <- data2 %>%
  select(-name) %>%
  left_join(plastic_data_all, by = "code")

# Member 1 (pair 1) write code to join your data to 'plastic_data_all':
plastic_data_all <- data3 %>%
  select(-name) %>%
  right_join(plastic_data_all, by = "code")

# Member 3 (pair 2) write code to join your data to 'plastic_data_all':
plastic_data_all <- data4 %>%
  select(-name) %>%
  right_join(plastic_data_all, by = "code")
```


```{r print-data, message = FALSE}
# Member 2 (pair 1) adds code here to view the first 10 rows of 'plastic_data_all'
plastic_data_all %>% head(n = 10)
```



## Exercises

### Ex 1.

```{r Ex1, message = FALSE}
# Write your code here to answer exercise 1
plastic_data_all %>% 
  count(region, sort = TRUE)

```



### Ex 2.

```{r Ex2, message = FALSE}
# Write your code here to answer exercise 2

plastic_data_all <- plastic_data_all %>%
  mutate(total_mismanaged_plastic = mismanaged_plastic * population)

plastic_data_all %>% 
  group_by(region) %>%
  summarise(avg_total = mean(total_mismanaged_plastic))

```


### Ex 3.

```{r Ex3_pt1, message = FALSE}
# Write your code here to answer exercise 3
plastic_data_all <- plastic_data_all %>%
  mutate(pct_mismanaged_plastic_ocean = mismanaged_plastic_ocean / mismanaged_plastic)

plastic_data_all %>% 
  group_by(region) %>%
  summarise(med_pct = median(pct_mismanaged_plastic_ocean))

plastic_data_all %>% filter(code == "SOM")
```

`NaN` stands for **Not a Number**. This has happened because both mismanaged plastic waste per capita estimates for Somalia is `0`, therefore the percentage calculation is attempting to compute `0/0`! As there is a missing value percentage value for Somalia, then it is not possible to compute the median statistic for Africa, hence why the summary table contains `NA`.

```{r Ex3_pt2, message = FALSE}
plastic_data_all %>% 
  drop_na() %>%
  group_by(region) %>%
  summarise(med_pct = median(pct_mismanaged_plastic_ocean)) %>%
  arrange(med_pct)
```



### Ex 4.

```{r Ex4, message = FALSE}
# Write your code here to answer exercise 4

plastic_data_all <- plastic_data_all %>%
  rename(
    pop = population,
    ocean_plastic = mismanaged_plastic_ocean,
    mismanaged = mismanaged_plastic,
    tot_plastic = total_mismanaged_plastic,
    pct_ocean = pct_mismanaged_plastic_ocean
  )


plastic_data_all <- plastic_data_all %>%
  mutate(region = str_replace_all(region, "Latin America and The Caribbean", "LAC"))

plastic_data_all %>% head(n = 10)

```

Create your own table/summary using the edited `plastic_data_all` data frame.