---
title: "Lab 06 - Conveying the right message through visualisation"
author: "KEY"
date: "`r Sys.Date()`"
output: html_document

---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(lubridate)
```


### Exercise 1

Values are approximate, and read from the graph.  (In my opinion this is easier to do with "mask" first, then "no mask", to avoid switching between axes.)

```{r create-data}
df <- tribble(
  ~date,       ~value, ~mask,
  "7/12/2020", 25.2,   "Mask",
  "7/13/2020", 19.8,   "Mask",
  "7/14/2020", 19.7,   "Mask",
  "7/15/2020", 20.5,   "Mask",
  "7/16/2020", 19.8,   "Mask",
  "7/17/2020", 19.8,   "Mask",
  "7/18/2020", 20.5,   "Mask",
  "7/19/2020", 19.8,   "Mask",
  "7/20/2020", 20.7,   "Mask",
  "7/21/2020", 21.4,   "Mask",
  "7/22/2020", 19.7,   "Mask",
  "7/23/2020", 19.7,   "Mask",
  "7/24/2020", 20.4,   "Mask",
  "7/25/2020", 19.1,   "Mask",
  "7/26/2020", 19.6,   "Mask",
  "7/27/2020", 17,     "Mask",
  "7/28/2020", 16.3,   "Mask",
  "7/29/2020", 16.5,   "Mask",
  "7/30/2020", 16.6,   "Mask",
  "7/31/2020", 16,     "Mask",
  "8/1/2020",  16.1,   "Mask",
  "8/2/2020",  15.8,   "Mask",
  "8/3/2020",  15.9,   "Mask",
  "7/12/2020", 9.8,    "No Mask",
  "7/13/2020", 9,      "No Mask",
  "7/14/2020", 9.3,    "No Mask",
  "7/15/2020", 9.8,    "No Mask",
  "7/16/2020", 9.9,    "No Mask",
  "7/17/2020", 9.4,    "No Mask",
  "7/18/2020", 9.4,    "No Mask",
  "7/19/2020", 9,      "No Mask",
  "7/20/2020", 8.4,    "No Mask",
  "7/21/2020", 8.4,    "No Mask",
  "7/22/2020", 8.6,    "No Mask",
  "7/23/2020", 8.4,    "No Mask",
  "7/24/2020", 9.8,    "No Mask",
  "7/25/2020", 9.9,    "No Mask",
  "7/26/2020", 10.1,   "No Mask",
  "7/27/2020", 9.5,    "No Mask",
  "7/28/2020", 9.5,    "No Mask",
  "7/29/2020", 9.6,    "No Mask",
  "7/30/2020", 10,     "No Mask",
  "7/31/2020", 8.8,    "No Mask",
  "8/1/2020",  9,      "No Mask",
  "8/2/2020",  8.8,    "No Mask",
  "8/3/2020",  9.1,    "No Mask"
)
```


### Exercise 2

```{r better-viz}
df %>%
  mutate(date = mdy(date)) %>%
  ggplot(aes(x = date, y = value, color = mask, group = mask)) +
  geom_line() +
  scale_color_manual(values = c("orange", "blue")) +
  theme_minimal() +
  labs(
    title = "These should approximately match the original graph...",
    subtitle = "7-day rolling average of daily cases per 100k population, Kansas\nMask counties vs. no-mask counties",
    x = "Date",
    y = "Case rate",
    color = "County category"
  )
```


### Exercise 3

The actual values are easier to compare since the $y$-axis is shared, and it's clearer that in no-mask counties there are overall fewer cases.  The two $y$-axes downplayed this in the original graph.

### Exercise 4

*Answers may vary...*

At face value, this suggests a correlation between case rates and whether or not counties enforced a mask mandate: counties with such a mandate on average seem to have more cases.  The case rate seems to be decreasing overall over the period of study in mask counties, whereas it seems fairly stable in no-mask counties.

What does this tell us?  Well, a supporter of masks might point to the decrease visible in the mask counties, and an opponent might suggest the lower case rates in no-mask counties suggest masks aren't all that useful---for either individual, the graph could be taken to agree or to conflict with their prior ideas about mask-wearing.

In reality, the graph on its own doesn't tell us very much.  We have no idea when mask mandates were introduced, so it's not clear that mask wearing is what triggered the falling case rate in this case.  On the other hand, we don't know anything about other demographic features of the counties in question: for instance, mask mandates are more likely to have been triggered in cities, and at this stage in the pandemic the disease was far more present in urban centres than rural ones.  (Indeed, any causation may have been the other way around: mask mandates may have been introduced *because* case rates were higher.)

The graph as originally presented was designed to reinforce the message of falling cases and play down the absolute case rates.  This is a mask-support message, and one you may well agree with!  But as a data scientist you have a responsibility not to mislead, even if you're doing so for a good cause.