---
title: "HW 03 - Money in politics"
author: "KEYS"
date: "`r Sys.Date()`"
output: html_document
---

### Load packages

```{r load-packages, message = FALSE}
library(tidyverse)
library(scales)
```

### Exercise 1

_Question_:

Write a function using a `for` loop in R to create a single data set and called it `pac_all` containing the contributions in all election years available. In your R Markdown file, load `pac_all.csv` and report its number of observations and variables as a full sentence.

_Answer_:

#### Code

```{r load-pac-all, message = FALSE}
dataset <- function(index){
  pac_aux <- NULL
  for (i in index){
    file_name <- paste("data/pac_", i, ".csv", sep = "")
    pac_aux <- rbind( pac_aux, read_csv(file_name) )
  }
  return(pac_aux)
}

years <- seq(from = 2000, to = 2020, by = 2)
pac_all <- dataset(years)
glimpse(pac_all)
```
Any code that merges the data correctly is acceptable as long as it contains a for loop. 

### Narrative

`pac-all.csv` has `r nrow(pac_all)` rows of observations corresponding to the contributions of foreign PACS for each year. There are `r ncol(pac_all)` variables in this dataset.

### Exercise 2

_Question_:

Use the `separate()` function to separate `country_parent` into `country` and `parent` columns. End your code chunk by printing out the top 10 rows of your data frame (if you just type the data frame name it should automatically do this for you).

_Answer_:

#### Code

**NOTE:** "Top 10" was intended to mean the first ten rows.  Do not penalise if the student has provided (e.g.) the ten rows with the highest values of some column.

```{r separate-name-country}
pac_all <- pac_all %>%
  separate(country_parent, 
           into = c("country", "parent"), 
           sep = "/", 
           extra = "merge")
```

```{r show-result-1}
pac_all %>% print(n = 10)
```

### Exercise 3

_Question_:

Remove the character strings including `$` and `,` signs in the `total`, `dems`, and `repubs` columns and convert these columns to numeric. End your code chunk by printing out the top 10 rows of your data frame (if you just type the data frame name it should automatically do this for you).

_Answer_:

#### Code

**NOTE:** Also ok if students do this without writing an additional `parse_currency` or similar function.

The function `parse_number` can *almost* deal with this on its own. The problem is that there are instances of negative dollar amounts (e.g. `"-$3,000"`) which it cannot handle (it removes commas anywhere and any non-numeric characters at the start, but because it treats the minus sign as a numeric character it then does not know what to do about the dollar sign).  This does mean that it is sufficient to remove the `$`s and then run `parse_number`, and students should receive full credit for this.

```{r clean-currency}
parse_currency <- function(x){
  x %>%
    # remove dollar sign
    str_remove("\\$") %>%
    # remove all occurrences of commas
    str_remove_all(",") %>%
    # convert to numeric
    as.numeric()
}

pac_all <- pac_all %>%
  mutate(
    total = parse_currency(total),
    dems = parse_currency(dems),
    repubs = parse_currency(repubs)
  )
```

```{r show-result-2}
pac_all %>% print(n = 10)
```

### Exercise 4

_Question_: Create a line plot of total contributions from all foreign-connected PACs in the Canada and Mexico over the years. Once you have made the plot, write a brief interpretation of what the graph reveals.

_Answer_:

#### Code

```{r ca-mex-viz}
pac_all %>%
  filter(country == "Canada" | country == "Mexico") %>%
  group_by(country, year) %>%
  summarise(total = sum(total),
            .groups = "drop") %>%
  
  ggplot( mapping = aes(x = year, 
                        y = total/1000000, 
                        colour = country) ) +
  geom_line(size = 1) +
  labs(
    title = "Total Contributions from All Foreign-connected PACs",
    subtitle = "in Canada and Mexico",
    x = "Year",
    y = "Total Contribution in million USD",
    colour = "Country") +
  theme_minimal()
```

#### Narrative

*(Students do not need to write quite this much for full credit!)*

As seen in the plot above, Canada's total contribution from all foreign-connected PACs is larger than PACs from Mexico, and has increased at a higher rate over time. Specifically, an increase in total PAC contribution spending from 2005 can be observed in both Canadian and Mexican PACs, however this spending increase is at a much slower rate for Mexican PACS. A potential factor for this larger increase in spending is that as of 2019 Canada have 14 companies in the Fortune Global 500 whereas Mexico only 4. This could be further investigated by looking at the number of companies, and size of contribution, that are made from the PACS in each respective country.

PACs in both countries appear to recently have decreased spending, with this occurring since 2016 in Mexico and 2018 in Canada. Although several factors likely have contributed to this decrease, one potential factor is the election of American president Donald Trump in 2016; who's foreign policy focused on an "America First" approach to immigration control and trade.

### Exercise 5

_Question_: Make this plot as pretty as possible by changing colours, background colour, fonts, or anything else you can think of.

_Answer_:

The plot should contain:

* Clear labels, scales and name of plot describing what is being plotted. (1 point)
* Edit theme(), to change colours and font to something more reasonable. (2 points)
* One point for creativity, the plot looks different than other plots used in the class (1 point).

For reference, the code for the ugly plot is,
```{r, echo=TRUE}
ggplot(data = mpg, 
       mapping = aes(x = cty, y = hwy, colour = drv)) +
  geom_point(alpha = 10) + 
  theme(plot.background = element_rect(fill = "green"),
        axis.line = element_line(size = rel(6), 
                                 arrow = arrow(), 
                                 colour = "black"),
        panel.background = element_rect(fill = "red", 
                                        colour = "grey50"),
        axis.title.y = element_text(size = rel(2.0), 
                                    angle = 99),
        axis.title.x = element_text(size = rel(.6), 
                                    angle = 40),
        panel.grid.major = element_line(colour = "green") )
```

