---
title: "hw-02-key"
author: "Your Name (S000000)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cherryblossom)
```



## Exercise 1

#### Question

The code below produces a simple histogram for the runners of the `"10 Mile"` event. 

```{r ex1, warning = FALSE, out.width="80%"}
run17 %>%
  filter(event == "10 Mile") %>%
  ggplot(mapping = aes(x = age)) +
    geom_histogram(binwidth = 2)
```

* Describe the overall _shape_ of the histogram. What does this suggest about the _structure_ of the age distribution of the runners? You may calculate some simple summary statistics to support your comments.
* Create a similar data visualisation for the `"5K"` race event. Does the age distribution look for the `"10 Mile"` and `"5K"` events appear to be similar or different? Justify your answer?

#### Answer

```{r ex1_summary}
run17 %>%
  group_by(event) %>%
  summarise(avg = mean(age, na.rm = TRUE),
            med = median(age, na.rm = TRUE),
            min = min(age, na.rm = TRUE), 
            max = max(age, na.rm = TRUE))
```


The shape of the age histogram increases sharply around the age of 25, peaks at about 30 years old and then steadily decreases towards the oldest runner of 10 Mile. This shape illustrates that the age distribution is _right skewed_, specifically the longer tail from the peak in the histogram is that which extends out towards the _right_ in the _positive_ direction. The _mean_ age is 37 years, the _median_ age is 35 years and the _mode_ age (from the histogram) is about 29 years. Since the data is right skewed, then we see that the mode is the smallest value, followed by the median and the mean value is the largest.



```{r ex1_5Khist, warning = FALSE, out.width="80%"}
run17 %>%
  filter(event == "5K") %>%
  drop_na() %>%
  ggplot(mapping = aes(x = age)) +
    geom_histogram(binwidth = 2)
```

The mean age of the runners of the 5K race is 38 years old, with median 36 years and mode (from the histogram) is about 32 years. Again we notice the relationship _mode_ $<$ _median_ $<$ _mean_, which is indicative of a _right/positively skewed_ distribution. However, range in these values (5K: 38-32=6, 10 Mile: 37-29=8) suggest that the age distribution for the 10 Mile race is more skewed than that for the 5K race. Furthermore, the histogram for the 5K race demonstrates a number of other features, such as a larger number of runners of a lower age with a minor peak around 12 years old.


## Exercise 2

#### Question

Create a data visualisation based on the description.

Make a sequence of boxplots for the time (in minutes) that it took the runners to complete the Cherryblossom race (after accounting for the staggered start). The boxplots should be orientated vertically (i.e. side-by-side) based on the runner's identified gender. Furthermore, the data visualisation should consist of two panels for the two different race distances, with independent axes. Finally, add appropriate text to the image in order to assist the reader in understanding the data visualisation.


Provide a brief comment about the relationships seen in the data visualisation created by the above description.


#### Answer

The data visualisation was generated using the following code:

```{r ex2, warning = FALSE, out.width="80%"}
 run17 %>% mutate(
   net_min = net_sec/60,
   sex = case_when(
     sex == "M" ~ "Men's Race",
     sex == "F" ~ "Women's Race"
   )) %>%
   drop_na() %>%
   ggplot(mapping = aes(y = net_min, x = sex)) +
   geom_boxplot() + 
   facet_wrap(~event, scales = "free") +
   labs(
     x = "",
     y = "Race time, in minutes",
     title = "Boxplot of the Cherryblossom Run, 2017"
   )
```

Note that it is possible to obtain the same data visualisation via some variation in coding for the filtering, mutating and graphical layers. Where appropriate, credit will be awarded where the code is constructive towards the re-creation of the data visualisation.


Comment: The times for the 5K event are noticably quicker thant the times for the 10 Mile event. Within each event, the range in times for participants in the Men's race and the Women's race are fairly similar, but the median estimates (the bar through the box) are different, with the median time of the Men's race being faster than that from the Women's race.



## Exercise 3

#### Question

Recreate the data visualisation.

#### Answer

```{r ex3, message = FALSE, out.width="80%"}
run17_edit <- run17 %>% 
  filter(event == "10 Mile") %>%
  mutate(
    wave = case_when(
      clock_sec - net_sec < 3*60 ~ "First Wave",
      clock_sec - net_sec > 23*60 ~ "Last Wave",
      TRUE ~ "Other Waves"), 
    net_min = net_sec/60  ) %>%
  filter( wave != "Other Waves" )
 
ggplot(data = run17_edit,
        mapping = aes(x = age, 
                      y = net_min, 
                      colour = wave))+
   geom_point(alpha = 0.5) +
   geom_smooth(method = "lm", se = FALSE) + 
   scale_colour_manual(values = c(
     "First Wave" = "orange",
     "Last Wave" = "forestgreen")) + 
   theme_bw() + 
   theme(legend.position = "bottom") +
   labs(
     x = "Runner's Age, in years",
     y = "Race Time, in minutes",
     title = "Scatter Plot of 10 Mile Run, Time Against Age",
     subtitle = "First vs Last Wave",
     colour = ""
   )
```


## Exercise 4

#### Question

Find two visualisations:

* One that you think provides a _good_ representation of the data, and
* One that you think provides a _bad_ representation of the data.

For each case, give _two_ reasons why you consider the data visualisation you have found are good/bad. Your reasons should make reference to the '4 respects' of data visualisations.

#### Answer

This is an open question to search for a _good_ and _bad_ data visualisation. As such the reasons and justifications for why the data visualisation is good or bad is highly dependent on the specific of the chosen image. In either case, emphasis should be given in relation to the 4 respects of _people_, _mathematics_, _data_ and _computers_.

Reasons why a data visualisation may be considered as being _good_ include:

-   The use of an appropriate visualisation style (box-plot, histogram, scatter plot, etc.) for the variable types.
-   Informative use of axis labels and title
-   Good use of additional visualisation techniques (colour, point shape, faceting) to add extra variables that does not overload the image.
-   The data visualisation is clean and the intended reader can immediately understand the structure of the data.
-   Visualisation contains design elements that are inclusive (e.g. colour blindness)


Reasons why a data visualisation may be considered as being _bad_ include:

-   There is a geometric flaw in the presentation of the data (e.g., bar heights not consistent with the data)
-   An overemphasis on making the visualisation 'pretty' at the sacrifice of the visualisation being informative
-   An ambiguity in how to understand the visualisation (e.g. two point types in the legend with the same appearance that describe different scenarios)
-   Overloading the visualisation to incorporate too many variables
-   Poor demonstration of basic mathematics (e.g. percentages not adding to 100%)
-   No title or axis label, or axis labels that are incomprehensible
-   Axes containing no values or present values on an inconsistent scale (e.g. values of 10, 20, 75 and 80 appearing at equal distances along the axis)
-   The visualisation has design elements that may inadvertently emphasise stereotypes. 








