---
title: "Iteration"
subtitle: "<br><br> Introduction to Data Science"
author: "University of Edinburgh"
date: "<br> 2024/2025"
output:
  xaringan::moon_reader:
    css: ["./xaringan-themer.css", "./slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---



```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
library(DT)
library(xaringanExtra)
xaringanExtra::use_panelset()
```

class: middle

# Iterations

- For loop
- While loop
- Mapping

---


## For loops in R

- The `for` loop iterates over each element in **values** (which can be a vector, list, or any other iterable object), executing the specified block of code once for each element.
- It is an entry controlled loop, i.e. first the expression in **values** is evaluated, 
then the **body of the loop is executed** for each of the values. 

For loop Syntax in R:
```{r,eval=FALSE}
for (<index> in <values>) {
  
  # do something 
  
}
```
---

## Example 1 - with a sequence

.pull-left[
```{r,echo=TRUE}
for (i in 1:10){
  
  print(5*i)
  
}
```
]
.pull-right[
Narrative:
- START. 
- Set the loop values as the sequence<br> `1 2 3 4 ... 10`
- Set index `i` to the 1st value, `1`
- Compute `5*i` and print.
- Set `index` to the next value, `2`
- Compute `5*i` and print.
- Set `index` to the next value, `3`
- ... (continue) ...
- Set `index` to the next value, `10`
- Compute `5*i` and print.
- Reached end of the values. 
- END.
]
---

## Example 2 - with a general vector

```{r,echo=TRUE}
# Create a vector with the days of the week
week <- c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 
          'Thursday', 'Friday', 'Saturday')

# using for loop to iterate over the values in the vector
for (day in week) {
  # displaying each string in the vector
  print(day)
}
```

---

## Generalising - `seq_len(n)`

.pull-left[
* Creates a sequence from `1` to `n`.
* An alternative method to `1:n`.
* Example:
```{r}
seq_len(6)   # same as 1:6
```

```{r}
n <- 5
for(i in seq_len(n)){
  print(LETTERS[i])
}
```

]

.pull-right[
- Useful for catching unexpected behaviour
```{r}
seq_len(0)  # an empty vector
```

```{r}
# Bad example
n <- 0
for(i in 1:0) print(LETTERS[i])

# Good example
n <- 0
for(i in seq_len(n)) print(LETTERS[i])
```

]

---

## Generalising - `seq_along(x)`

.pull-left[
* Creates an index sequence based on the input `x`

* Example:
```{r}
week <- c('Sun', 'Mon', 'Tues', 'Wed', 
          'Thurs', 'Fri', 'Sat')
seq_along(week)
```
]

.pull-right[
```{r}
week <- c('Sun', 'Mon', 'Tues', 'Wed', 
          'Thurs', 'Fri', 'Sat')
#Compute number of characters in each string

#Set-up blank vector to store calculations
num_char <- rep(0L, length(week))
for(i in seq_along(week)){
  #Do calc. for i-th weekday
  num_char[i] <- nchar(week[i])
}
num_char
```

```{r}
nchar(week)
```
]

---

class: middle

# While loop



---

## While loops in R


- The `while` loop in R repeatedly executes a block of code as long as the specified **condition** evaluates to `TRUE`.
- In this loop, the **test condition** is evaluated at the beginning of each iteration; if it evaluates to `TRUE`, the **body of the loop** is executed.

While loop Syntax in R:
```{r,eval=FALSE}
while (<condition>) {
  
  #statement
  
}
```
---

## Example 1 - for loop as while loop

.pull-left[
```{r, echo=TRUE}
i <- 0
while (i < 10){
  print(5 * i)
  i <- i + 1 # IMPORTANT!
}
```
]
.pull-right[
Narrative:
* Initialise `i` with `0`
* Enter while loop
* TEST: `i < 10`, i.e. `0 < 10` -> `TRUE`
* Compute `5*i` and print
* Add `1` to `i` & re-assign (`i == 1`)
* TEST: `i < 10`, i.e. `1 < 10` -> `TRUE`
* Compute `5*i` and print
* Add `1` to `i` & re-assign (`i == 2`)
* ... (continue) ...
* TEST: `i < 10`, i.e. `10 < 10` -> `FALSE`
* Exit while loop
]

---

## Caution! Infinite loops


.pull-left[
```{r, echo=TRUE , eval = FALSE}
i <- 0
while (i < 10){
  print(5 * i)
  i <- i - 1  #<<
}
```

```{r echo = FALSE}
for(i in 0:-20){
  print(5*i)
}
```

]
.pull-right[
Narrative:
* Initialise `i` with `0`
* Enter while loop
* TEST: `i < 10`, i.e. `0 < 10` -> `TRUE`
* Compute `5*i` and print
* Update `i` (`i = -1`)
* TEST: `i < 10`, i.e. `-1 < 10` -> `TRUE`
* Compute `5*i` and print
* Update `i` (`i = -2`)
* ... (continue) ...
* ... (FOR EVER!) ...
* .small[... (AND EVER!) ...]
* .small[... (AND EVER!) ...]
* .tiny[... (AND EVER!) ...]
* .tiny[... (AND EVER!) ...]
]

---
## Main differences between for and while loop.

- A `for` loop is used when the number of iterations a code should be run is **known**;
a `while` loop is used when the number of iterations is **not known**.

- `while` loops can potentially result in infinite loops if not written properly.

- In `for` loops the evaluation of the iteration <tt>values</tt> is done once and never repeated.

- In `while` loops the test condition is evaluated each time the loop iterates.

- If you accidental create an infinite `while` loop, either:
  - Press the `Esc` key to stop all computation, or
  - Click on the stop button, ðŸŸ¥ (for code chunk) or ðŸ›‘ (for console/viewer panels)

---

## Example 2 - convergence of algorithm

.panelset[
.panel[.panel-name[Gradient descent algorithm]
Use gradient descent algorithm to minimize a function. 

- Initialize the starting point $x_1$ 
- Set the learning rate $\alpha>0$
- $x_{n+1} = x_n - \alpha \cdot f'(x_n)$
- Stop when $\vert x_{n+1}-x_n \vert < \epsilon$

For example, let's find the minimum of $f(x) = x^2 -4x+4$.
]

.panel[.panel-name[Code]
```{r grad, results='hide'}
f <- function(x) { x^2 - 4*x + 4 }
df <- function(x) { 2*x - 4 }
epsilon <- 1e-5   # Convergence criterion (tolerance)
alpha <- 0.3      # Learning rate
x <- 3            # Initial guess
error <- 1        # For initial step
iter <- 1

while(error > epsilon) {  # Gradient descent algorithm
  gradient <- df(x)
  new_x <- x - alpha * gradient
  error <- abs(new_x - x) # Update error
  x <- new_x              # Update the guess
  cat("Iteration", iter, ": x =", x, "error =", error, "f(x) =", f(x), "\n")
  iter <- iter + 1        # Update the iteration
}

cat("Minimum occurs at x =", x, "\n")
```
]

.panel[.panel-name[Output]
```{r ref.label = "grad", echo = FALSE, warning = FALSE}
```
]
]

---

## Example 2 - convergence of algorithm

Note that I used 
```{r eval = FALSE}
while(error > epsilon){
  # ...
}
```

instead of 
```{r eval = FALSE}
while(error != 0){
  # ...
}
```

Why? 

--

Floating point calculations!

---

## Unexpected behavour with floating point numbers

.pull-left[
```{r, echo=TRUE}
#EXAMPLE 1
2/7 + 3/7 + 2/7
2/7 + 3/7 + 2/7 == 1
2/7 + 3/7 + 2/7 - 1
```
]
--
.pull-right[
```{r, echo=TRUE}
#EXAMPLE 2
1 + 10^(-16) == 1
10^(-16) 
.Machine$double.eps
```
]


---

## Solutions: use a **tolerance** value, or `near`.

```{r}
#EXAMPLE 1
2/7 + 3/7 + 2/7 == 1
```

.pull-left[
```{r, echo=TRUE}
tolerance <- 10^-16  # Something small

2/7 + 3/7 + 2/7 - 1 < tolerance
```
]
--
.pull-right[
```{r, echo=TRUE}
# default tol = .Machine$double.eps^0.5
# which approx is 1.5e-8 
near(2/7 + 3/7 + 2/7,1)
```
]

---

## `next` and `break` statements

- The `next` statement terminates the current iteration early and immediately proceeds to the next iteration of the loop.
- The `break` statement exits the entire loop prematurely, halting any further iterations.

.pull-left[
```{r}
for(i in 1:6){
  if(i %% 2 == 0){
    next
  }
  print(i)
}
```
]
.pull-right[
```{r}
i <- 1
while(i < 6){
  if(i == 3){
    break
  }
  print(i)
  i <- i + 1
}
```
]

---

class: middle

# Mapping: 
## To loop or not to loop, that is the question! 

---

## Avoiding Explicit Loops 

Suppose we have exam 1 and exam 2 scores of 10 students stored in a list...

```{r}
exam_scores <- list(
  exam1 <- c(80, 90, 70, 50, 80, 40, 90, 95, 80, 90),
  exam2 <- c(85, 83, 45, 60, 85, 95, 40, 75, 90, 85)
)
```

Calculate the mean of exam 1 and exam 2 using a loop:

```{r, echo=TRUE}

m <- numeric(length(exam_scores))
for (i in seq_along(exam_scores)){
  m[i] <- mean(exam_scores[[i]])
}
m

```

Can we make the R code both simpler and more efficient?


---

## `map_something`

Functions for looping over an object and returning a value (of a specific type):

* `map()` - returns a list
* `map_lgl()` - returns a logical vector
* `map_int()` - returns a integer vector
* `map_dbl()` - returns a double vector
* `map_chr()` - returns a character vector
* `map_df()` / `map_dfr()` - returns a data frame by row binding
* `map_dfc()` - returns a data frame by column binding
* ...

---

## How does mapping work?

Suppose we have `exam1` and `exam2` scores of 10 students stored in a list...

```{r}
exam_scores <- list(
  exam1 <- c(80, 90, 70, 50, 80, 40, 90, 95, 80, 90),
  exam2 <- c(85, 83, 45, 60, 85, 95, 40, 75, 90, 85)
)
```


...and we find the mean score in each exam

.pull-left[
```{r}
map(exam_scores, mean)
```
]
.pull-right[
```{r}
map_dbl(exam_scores, mean)
```
]
---

## Comparing execution times â±

.pull-left[
* Via looping
```{r,echo=TRUE}
start.time <- Sys.time()
m <- numeric(length(exam_scores))
for (i in seq_along(exam_scores)){
  m[i] <- mean(exam_scores[[i]])
}
end.time <- Sys.time()
m
end.time - start.time
```
]
.pull-right[
* Via mapping
```{r,echo=TRUE}
start.time <- Sys.time()
out <- map_dbl(exam_scores, mean)
end.time <- Sys.time()
out
end.time - start.time
```
]
---

## Examples

```{r}
map_dbl(mtcars, median)
```


```{r}
map_lgl(mtcars,~any(is.na(.x)))
```

```{r}
week <- c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 
          'Thursday', 'Friday', 'Saturday')
map_chr(week, \(word) paste(word,"has",nchar(word),"letters."))
```


---

## Chat-GPT, give examples of `for` loops

I asked Chat-GPT (actually, ELM) to give me examples of uses of for-loops in R for students taking an Introduction to Data Science class.

.panelset[
.panel[.panel-name[Example 1]
Summarize each column of a dataset
.pull-left[
```{r eval = FALSE}
mean_values <- numeric(ncol(mtcars))
for (i in 1:ncol(mtcars)) {
  mean_values[i] <- mean(mtcars[[i]])
}
```
]
.pull-right[
```{r eval = FALSE}
mtcars %>% 
  summarise(across(everything(), mean))
```
]
]
.panel[.panel-name[Example 2]
Modify the value of a variable
.pull-left[
```{r eval = FALSE}
for (i in 1:nrow(mtcars)) {
  if(mtcars$am[i] == 1){
    mtcars$am[i] <- "manual"
  } else {
    mtcars$am[i] <- "automatic"
  }
}
```
]
.pull-right[
```{r eval = FALSE}
mtcars %>% 
  mutate(am=ifelse(am == 1,
                   "manual","automatic"))
```
]
]
.panel[.panel-name[Example 3]
Bootstrap (come back in w10L20!)
.pull-left[
```{r eval = FALSE}
n_bootstraps <- 1000
bootstrap_means <- numeric(n_bootstraps)
for (i in 1:n_bootstraps) {
  resample <- sample(mtcars$mpg, replace = TRUE)
  bootstrap_means[i] <- mean(resample)
}
```
]
.pull-right[
```{r eval = FALSE}
replicate(1000, mtcars$mpg %>% 
            sample(size = length(.), 
                   replace = TRUE) %>% 
            mean())
```
]
]
.panel[.panel-name[Example 4]
Processing multiple files
.pull-left[
```{r eval = FALSE}
# files <- ... # provide list of files
data_list <- list()
for (file in files) {
  dat <- read.csv(file)
  # user-defined function to process data
  cleaned <- clean_data(dat) 
  data_list[[file]] <- cleaned
}
```
]
.pull-right[
```{r eval = FALSE}
data_list <- map(files, 
                 ~ read_csv(.x) %>% 
                   clean_data)
```
]
]
]



