---
title: "Iteration"
subtitle: "<br><br> Introduction to Data Science"
author: "University of Edinburgh"
date: "<br> 2023/2024"
output:
  xaringan::moon_reader:
    css: ["./xaringan-themer.css", "./slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---



```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
library(DT)
```

class: middle

# Iterations

---
## For loops in R

- The `for` loop will iterate the expression through the number of components in `values` (list, matrix or vector) one at a time.
- It is an entry controlled loop, in this loop the **values** are tested first, then the **body of the loop is executed** in each of the values.
<br>
<br>
<br>
---
## For loops in R

- The `for` loop will iterate the expression through the number of components in `values` (list, matrix or vector) one at a time.
- It is an entry controlled loop, in this loop the **values** are tested first, then the **body of the loop is executed** in each of the values.

R – For loop Syntax:
```{r,eval=FALSE}
for (values) {
  
  #statement
  
}
```
---

#Example 1

```{r,echo=TRUE}
for (i in 1:10){
  
  print(5*i)

  }
```
---

#Example 2

```{r,echo=TRUE}
# Create a vector with the days of the week
week <- c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday',
           'Friday', 'Saturday')
 
# using for loop to iterate over the values in the vector
for (day in week)
{
     # displaying each string in the vector
    print(day)
}
```
---

## While loops in R

- The `while` loop will iterate the expression as long as the component in `condition` is TRUE.
- In this loop the **test condition** is tested first, then the **body of the loop is executed**, the loop body would not be executed if the test condition is false.
<br>
<br>
<br>
---
## While loops in R

- The `while` loop will iterate the expression as long as the component in `condition` is TRUE.
- In this loop the **test condition** is tested first, then the **body of the loop is executed**, the loop body would not be executed if the test condition is false.

R – While loop Syntax:
```{r,eval=FALSE}
while (condition) {
  
  #statement
  
}
```
---

#Example 1

```{r,echo=TRUE}
i <- 0

while (i < 10){

    print(5*i)
   i <- i+1  # Be careful or is an infinite loop

   }
```
---

#Example 2

Program to calculate factorial of a number.

.pull-left[
```{r,echo=TRUE}

# create variable whose 
# factorial will be calculated
n <- 5
# assigning the factorial variable
# and iteration variable to 1
factorial <- 1
i <- 1
 
# using while loop
while (i <= n)
{
 
    # multiplying the factorial variable
    # with the iteration variable
    factorial = factorial * i
 
    # add one iteration 
    i = i + 1
}
```
]
.pull-right[
```{r,echo=TRUE}

# displaying the factorial
print(factorial)
```
]
---
See the following example to illustrate how similiar for and while loops are:
.pull-left[
```{r, echo=TRUE, R.options=list(width = 40)}
values <- 1

while(values <= 10) {
        print(values)
        values <- values + 1
}
```
]
.pull-right[
```{r, echo=TRUE,R.options=list(width = 80)}
# this for loop provides the same output
values <- 10

for(i in 1:values) {
        print(i)
}

```
]

---
# Main differences between for and while loop.

- For loop is used when the number of iterations a code should be run is **known** where a while loop is used when the number of iterations is **not known**.

- `while` loops can potentially result in infinite loops if not written properly.

- In `for` loop the initialization once done is never repeated.

- In `while` loop the initialization is done each time the loop iterate.

---

## Repeat loops in R

- The `repeat` loop will run the same `statement` until the `stop condition` has been encountered. 
- If no stop condition is present in the body of the repeat loop then it will iterate infinitely.

<br>
<br>
<br>
---
## Repeat loops in R

- The `repeat` loop will run the same `statement` until the `stop condition` has been encountered. 
- If no stop condition is present in the body of the repeat loop then it will iterate infinitely.

R – Repeat loop Syntax:
```{r,eval=FALSE}
repeat 
{ 
   statement
 
   if( stop condition ) 
   {
      break
   }
}
```
---

#Example 1

```{r,echo=TRUE}
i <- 1

repeat
{
    # statements
    print(i)
    i <- i+ 1
 
    # Stop condition
    if(i > 5)
    {
        # using break statement
        # to terminate the loop
        break
    }
}
```
---

#Example 2

```{r,echo=TRUE}
# initializing the iteration variable with 0
i <- 0
repeat
{
    # statement to be executed multiple times
    print("Hello world")
 
    # incrementing the iteration variable
    i <- i + 1
 
    # checking the stop condition
    if (i == 5)
    {
        # terminate the loop
        break
    }
}
```
---
**Avoid** to create an infinite loop by mistake.

- Be careful with the condition being `TRUE` and `FALSE`.

<br>
<br>
<br>

---

**Avoid** to create an infinite loop by mistake.

- Be careful with the condition being `TRUE` and `FALSE`.

Examples

.pull-left[
```{r, echo=TRUE}
#EXAMPLE 1
2/7 + 3/7 + 2/7
2/7 + 3/7 + 2/7 == 1
2/7 + 3/7 + 2/7 - 1

```
]
.pull-right[
```{r, echo=TRUE}
#EXAMPLE 2
1 + 10^(-16) == 1
10^(-16) 
.Machine$double.eps
```
]
---

Instead use a **tolerance** value.


```{r, echo=TRUE}
#EXAMPLE 1
2/7 + 3/7 + 2/7 == 1

tolerance <- 10^-16  #Something you consider small. 

2/7 + 3/7 + 2/7 - 1 < tolerance

tolerance <- .Machine$double.eps #The machine epsilon

2/7 + 3/7 + 2/7 - 1 < tolerance

```
---

- R like other numerical software in existence, uses **floating point arithmetic** to do its calculations.

- Represents real numbers approximately, using an integer with a fixed precision, scaled by an integer exponent of a fixed base. 

- For example, $12.345$ can be represented as a base-ten floating-point number:
$$12345\times10^{-3}$$
- Be careful and instead use the **tolerance** value inside the `conditions`.

---

##Example

```{r, echo=TRUE}

j <- 0
while (j < 1) { j <- j+0.1 }; j
j <- 0
j + .1 + .1 + .1 + .1 + .1 + .1 + .1 + .1 + .1 + .1
j + .1 + .1 + .1 + .1 + .1 + .1 + .1 + .1 + .1 + .1 < 1

j <- 0
tolerance <- 10^-16 
while (j  < 1 - tolerance) { j <- j+0.1 }; j

```

---
## Avoiding Explicit Loops 

Suppose we have exam 1 and exam 2 scores of 10 students stored in a list...

```{r}
exam_scores <- list(
  exam1 <- c(80, 90, 70, 50, 80, 40, 90, 95, 80, 90),
  exam2 <- c(85, 83, 45, 60, 85, 95, 40, 75, 90, 85)
)
```

Calculate the mean of exam 1 and exam 2 using a loop:

```{r, echo=TRUE}

m <- numeric()
for (i in 1:length(exam_scores)){
  m[i] <- mean(exam_scores[[i]])
}
m

```
---

class: middle
# We want to make R code both simpler and more efficient.

---

## How does mapping work?

Suppose we have exam 1 and exam 2 scores of 10 students stored in a list...

```{r}
exam_scores <- list(
  exam1 <- c(80, 90, 70, 50, 80, 40, 90, 95, 80, 90),
  exam2 <- c(85, 83, 45, 60, 85, 95, 40, 75, 90, 85)
)
```

--

...and we find the mean score in each exam

```{r}
map(exam_scores, mean)
```

---

...and suppose we want the results as a numeric (double) vector

```{r}
map_dbl(exam_scores, mean)
```


---

## `map_something`

Functions for looping over an object and returning a value (of a specific type):

* `map()` - returns a list
* `map_lgl()` - returns a logical vector
* `map_int()` - returns a integer vector
* `map_dbl()` - returns a double vector
* `map_chr()` - returns a character vector
* `map_df()` / `map_dfr()` - returns a data frame by row binding
* `map_dfc()` - returns a data frame by column binding
* ...

---

Example of the students
```{r,echo=TRUE}
exam_scores <- list(
  exam1 <- c(80, 90, 70, 50, 80, 40, 90, 95, 80, 90),
  exam2 <- c(85, 83, 45, 60, 85, 95, 40, 75, 90, 85)
)
```
---
Compare the **times** of the two approaches.
.pull-left[
```{r,echo=TRUE}
start.time <- Sys.time()
m <- numeric()
for (i in 1:length(exam_scores)){
  m[i] <- mean(exam_scores[[i]])
}
m
end.time <- Sys.time()
end.time - start.time
```
]
.pull-right[
```{r,echo=TRUE}
start.time <- Sys.time()
map_dbl(exam_scores, mean)
end.time <- Sys.time()
end.time - start.time
```
]
---

Example of the students with 10 exams and 100 students.
```{r,echo=TRUE,R.options=list(max.print = 10)}
num_students <- 100
possible_marks <- seq(from = 0, to = 100, by = 1)
exam_scores <- list(
  exam1  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam2  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam3  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam4  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam5  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam6  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam7  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam8  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam9  <- sample(x = possible_marks, size = num_students, replace = TRUE),
  exam10 <- sample(x = possible_marks, size = num_students, replace = TRUE)
)
head(exam_scores)
```
---
Compare the **times** of the two approaches.
.pull-left[
```{r,echo=TRUE,R.options=list(max.print = 5)}
start.time <- Sys.time()
m <- numeric()
for (i in 1:length(exam_scores)){
  m[i] <- mean(exam_scores[[i]])
}
m
end.time <- Sys.time()
end.time - start.time
```
]
.pull-right[
```{r,echo=TRUE,R.options=list(max.print = 5)}
start.time <- Sys.time()
map_dbl(exam_scores, mean)
end.time <- Sys.time()
end.time - start.time
```
]
---
class: middle

# Useful functions to deal with strings.
---

## `str` family of functions on package `stringr`

```{r,echo=TRUE}
ls("package:stringr")
```

---

## Example
```{r,echo=TRUE,R.options=list(max.print = 5)}
#Create character vector
my_vector <- "Hey there everyone."

#remove first occurrence of "e" from vector
str_remove(my_vector, "e")

#remove all occurrences of "e" from vector
str_remove_all(my_vector, "e")

```

---

## The `separate` function

## Example

```{r, echo=TRUE}
#create data frame
df <- data.frame(player=c('A', 'A', 'B', 'B', 'C', 'C'),
                 year=c(1, 2, 1, 2, 1, 2),
                 stats=c('22-2', '29-3', '18-6', '11-8', '12-5', '19-2'))

df
```
---

## The `separate` function

## Example

```{r, echo=TRUE}

separate(df, col=stats, into=c('points', 'assists'), sep='-')
```
