---
title: "Model checking and multiple linear regression"
subtitle: "<br><br> Introduction to Data Science"
author: "University of Edinburgh"
date: "<br> 2022/23"
output:
  xaringan::moon_reader:
    css: ["./xaringan-themer.css", "./slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false

---


```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(magick)
library(dplyr)
library(tidymodels)
library(ggtext)
library(knitr)
library(kableExtra)
library(xaringanExtra)
library(Tmisc)
library(emo)
library(patchwork)
library(plotly)
library(widgetframe)

xaringanExtra::use_panelset()
```


```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE,
  dplyr.print_min = 6, 
  dplyr.print_max = 6,
  tibble.width = 65,
  width = 65
  )
# figure height, width, dpi
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 8, 
                      fig.asp = 0.618,
                      out.width = "60%",
                      fig.align = "center",
                      dpi = 300,
                      message = FALSE)
# ggplot2
ggplot2::theme_set(ggplot2::theme_gray(base_size = 16))
# set seed
set.seed(1234)
# fontawesome
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}
# output number of lines
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

```

layout: true
  
<div class="my-footer">
<span>
University of Edinburgh
</span>
</div> 

---
## Reminders

- HW03 due on Friday
- Look through lab sheet in advance of lab tomorrow
- Meet with your team to work on the project

---
## Topics

- Assessing model fit.
- Linearity and transformations.
- Multiple linear regression
- Model comparison
- Interaction effects


---
class: middle

# Model checking

---

## Data: Paris Paintings

```{r message=FALSE}
pp <- read_csv("data/paris-paintings.csv", na = c("n/a", "", "NA"))
```

- Number of observations: `r nrow(pp)`
- Number of variables: `r ncol(pp)`

---

## "Linear" models

- We're fitting a "linear" model, which assumes a linear relationship between our explanatory and response variables.
- But how do we assess this?

---

## Graphical diagnostic: residuals plot

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "residuals-plot", echo = FALSE, warning = FALSE, out.width = "60%"}
```
]
.panel[.panel-name[Code]
```{r residuals-plot, fig.show="hide"}
ht_wt_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(Height_in ~ Width_in, data = pp)

ht_wt_fit_aug <- augment(ht_wt_fit$fit) #<<

ggplot(ht_wt_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted height", y = "Residuals")
```
]

.panel[.panel-name[Augment]
```{r}
ht_wt_fit_aug
```
]
]

---

## More on `augment()`

```{r}
glimpse(ht_wt_fit_aug)
```


---

## Looking for...

- Residuals distributed randomly around 0
- With no visible pattern along the x or y axes

```{r out.width = "60%", echo=FALSE}
df <- tibble(
  fake_resid = rnorm(1000, mean = 0, sd = 30),
  fake_predicted = runif(1000, min = 0, max = 200)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Fan shapes**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_resid = c(rnorm(100, mean = 0, sd = 1), 
                 rnorm(100, mean = 0, sd = 15), 
                 rnorm(100, mean = 0, sd = 25), 
                 rnorm(100, mean = 0, sd = 20), 
                 rnorm(100, mean = 0, sd = 25), 
                 rnorm(100, mean = 0, sd = 50), 
                 rnorm(100, mean = 0, sd = 35), 
                 rnorm(100, mean = 0, sd = 40),
                 rnorm(200, mean = 0, sd = 80)),
  fake_predicted = seq(0.2, 200, 0.2)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Groups of patterns**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_predicted = seq(0.2, 200, 0.2),
  fake_resid = c(
    rnorm(500, mean = -20, sd = 10),
    rnorm(500, mean = 10, sd = 10)
  )
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Residuals correlated with predicted values**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_predicted = seq(0.2, 200, 0.2),
  fake_resid = fake_predicted + rnorm(1000, mean = 0, sd = 50)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

## Not looking for...

.large[
**Any patterns!**
]

```{r out.width = "60%", echo=FALSE}
set.seed(12346)
df <- tibble(
  fake_predicted = seq(-100, 100, 0.4),
  fake_resid = -5*fake_predicted^2 - 3*fake_predicted + 20000 + rnorm(501, mean = 0, sd = 10000)
)
ggplot(df, mapping = aes(x = fake_predicted, y = fake_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted", y = "Residuals")
```

---

.question[
What patterns does the residuals plot reveal that should make us question whether a linear model is a good fit for modeling the relationship between height and width of paintings?
]

```{r out.width = "60%", echo=FALSE}
ggplot(ht_wt_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted height", y = "Residuals")
```

---

class: middle

# Linearity and transformations

---

## Data: Paris Paintings

```{r echo=FALSE, out.width = "70%"}
ggplot(data = pp, aes(x = price)) +
  geom_histogram(binwidth = 1000) +
  labs(title = "Prices of paintings")
```

---

## Price vs. width

```{r echo=FALSE, out.width = "70%", warning = FALSE}
ggplot(data = pp, aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(x = "Width (in)", y = "Price (livres)")
```

---

## Focus on paintings with `Width_in < 100`

That is, paintings with width < 2.54 m

```{r}
pp_wt_lt_100 <- pp %>% 
  filter(Width_in < 100)
```

---

## Price vs. width

.question[
Which plot shows a more linear relationship?
]

.small[
  
.pull-left[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Price vs. width", 
    subtitle = "For width < 100 in",
    x = "Width (inches)", 
    y = "Price (livres)"
    )
```
]

.pull-right[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Log(price) vs. width", 
    subtitle = "For width < 100 in",
    x = "Width (inches)", 
    y = "Log(price) (log livres)"
    )
```
]

]

---

## Price vs. width, residuals

.question[
Which plot shows a residuals that are uncorrelated with predicted values from the model? Also, what is the unit of the residuals?
]
  
.pull-left[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
price_wt_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(price ~ Width_in, data = pp_wt_lt_100)
price_wt_fit_aug <- augment(price_wt_fit$fit)

ggplot(data = price_wt_fit_aug, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Price vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted price (livres)", 
    y = "Residuals"
    )
```
]
.pull-right[
```{r message=FALSE, echo=FALSE, out.width = "100%"}
pp_wt_lt_100 <- pp_wt_lt_100 %>% 
  mutate(log_price = log(price))

lprice_wt_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_price ~ Width_in, data = pp_wt_lt_100)
lprice_wt_fit_aug <- augment(lprice_wt_fit$fit)

ggplot(data = lprice_wt_fit_aug, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Log(Price) vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted log(price) (log livres)", 
    y = "Residuals"
    )
```
]

---

## Transforming the data

- We saw that `price` has a right-skewed distribution, and the relationship between price and width of painting is non-linear.

--
- In these situations a transformation applied to the response variable may be useful.

--
- In order to decide which transformation to use, we should examine the distribution of the response variable.

--
- The extremely right skewed distribution suggests that a log transformation may 
be useful.
    - log = natural log, $ln$
    - Default base of the `log` function in R is the natural log: <br>
    `log(x, base = exp(1))`
    
---

## Logged price vs. width

.question[
How do we interpret the slope of this model?
]

```{r echo=FALSE, message=FALSE, out.width="60%"}
ggplot(data = pp_wt_lt_100, mapping = aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "#8E2C90", se = FALSE) +
  labs(x = "Width (in)", y = "Log(price) (log livres)")
```

---

## Models with log transformation

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(log(price) ~ Width_in, data = pp_wt_lt_100) %>%
  tidy()
```

---


## Interpreting the slope

$$ \widehat{log(price)} = 4.67 + 0.0192 Width $$

--
- For each additional inch the painting is wider, the log price of the painting is expected to be higher, on average, by 0.0192 livres.

--
- which is not a very useful statement...

---

## Working with logs

- Subtraction and logs: $log(a) − log(b) = log(a / b)$

--
- Natural logarithm: $e^{log(x)} = x$

--
- We can use these identities to "undo" the log transformation

---

## Interpreting the slope

The slope coefficient for the log transformed model is 0.0192, meaning the log price difference between paintings whose widths are one inch apart is predicted to be 0.0192 log livres.

--

.question[
Using this information, and properties of logs that we just reviewed, fill in the blanks in the following alternate interpretation of the slope:

>For each additional inch the painting is wider, the price of the painting is expected to be `___` , on average, by a factor of `___`.
]


---

>For each additional inch the painting is wider, the price of the painting is expected to be `___` , on average, by a factor of `___`.


$$ log(\text{price for width x+1}) - log(\text{price for width x}) = 0.0192 $$

--

$$ log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right) = 0.0192 $$

--

$$ e^{log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right)} = e^{0.0192} $$

--

$$ \frac{\text{price for width x+1}}{\text{price for width x}} \approx 1.02 $$

--

For each additional inch the painting is wider, the price of the painting is expected to be higher, on average, by a factor of 1.02.

---

## Recap

- Non-constant variance is one of the most common model violations, however it is usually fixable by transforming the response (y) variable.

--
- The most common transformation when the response variable is right skewed is the log transform: $log(y)$, especially useful when the response variable is 
(extremely) right skewed.

--
- This transformation is also useful for variance stabilization.

--
- When using a log transformation on the response variable the interpretation of 
the slope changes: *"For each unit increase in x, y is expected on average to be higher/lower <br> by a factor of $e^{b_1}$."*

--
- Another useful transformation is the square root: $\sqrt{y}$, especially 
useful when the response variable is counts.

---

## Transform, or learn more?

- Data transformations may also be useful when the relationship is non-linear
- However in those cases a polynomial regression may be more appropriate
  + This is beyond the scope of this course, but you’re welcomed to try it for your final project, and I’d be happy to provide further guidance

---
class: middle

# The linear model with multiple predictors

---

## Data: Book weight and volume

The `allbacks` data frame gives measurements on the volume and weight of 15 books, some of which are paperback and some of which are hardback

.pull-left[
- Volume - cubic centimetres
- Area - square centimetres
- Weight - grams
]
.pull-right[
.small[
```{r echo=FALSE}
library(DAAG)
as_tibble(allbacks) %>%
  print(n = 15)
```
]
]

.footnote[
.small[
These books are from the bookshelf of J. H. Maindonald at Australian National University.
]
]

---

## Book weight vs. volume

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
ggplot(allbacks, aes(x = volume, y = weight)) +
  geom_point(alpha = 0.7, size = 3)
```
]

---

## Book weight vs. volume and cover

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
ggplot(allbacks, aes(x = volume, y = weight, color = cover, shape = cover)) +
  geom_point(alpha = 0.7, size = 3) +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#E48957", "#071381"))
```
]

---

## Interpretation of estimates

```{r echo=FALSE}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks) %>%
  tidy()
```

--
- **Slope - volume:** *All else held constant*, for each additional cubic centimetre books are larger in volume, we would expect the weight to be higher, on average, by 0.718 grams.

--
- **Slope - cover:** *All else held constant*, paperback books weigh, on average, 184 grams less than hardcover books.

--
- **Intercept:** Hardcover books with 0 volume are expected to weigh 198 grams, on average. (Doesn't make sense in context.)


---

## Main vs. interaction effects

.question[
Suppose we want to predict weight of books from their volume and cover type 
(hardback vs. paperback). Do you think a model with main effects or 
interaction effects is more appropriate? Explain your reasoning.

**Hint:** Main effects would mean rate at which weight changes as volume 
increases would be the same for hardback and paperback books and interaction 
effects would mean the rate at which weight changes as volume 
increases would be different for hardback and paperback books.
]

---

```{r book-main-int, echo=FALSE, out.width="65%", fig.asp = 0.8}
book_main_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks)
book_main_fit_aug <- augment(book_main_fit$fit)

book_int_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover + volume*cover, data = allbacks)
book_int_fit_aug <- augment(book_int_fit$fit)

p_main <- ggplot() +
  geom_point(book_main_fit_aug, 
             mapping = aes(x = volume, y = weight, color = cover, shape = cover), alpha = 0.7) +
  geom_line(book_main_fit_aug, 
            mapping = aes(x = volume, y = .fitted, color = cover)) +
  labs(title = "Main effects, parallel slopes", 
       subtitle = "weight-hat = volume + cover") +
  scale_color_manual(values = c("#E48957", "#071381"))

p_int <- ggplot() +
  geom_point(book_int_fit_aug, 
             mapping = aes(x = volume, y = weight, color = cover, shape = cover), alpha = 0.7) +
  geom_line(book_int_fit_aug, 
            mapping = aes(x = volume, y = .fitted, color = cover)) +
  labs(title = "Interaction effects, not parallel slopes", 
       subtitle = "weight-hat = volume + cover + volume * cover") +
  scale_color_manual(values = c("#E48957", "#071381"))

p_main /  p_int  + plot_layout(guides = "collect")
```

---

## In pursuit of Occam's razor

- Occam's Razor states that among competing hypotheses that predict equally well, the one with the fewest assumptions should be selected.

--
- Model selection follows this principle.

--
- We only want to add another variable to the model if the addition of that variable brings something valuable in terms of predictive power to the model.

--
- In other words, we prefer the simplest best model, i.e. **parsimonious** model.

---

.pull-left[
```{r ref.label = "book-main-int", echo = FALSE, out.width="100%", fig.asp = 0.8}
```
]
.pull-right[
.question[
Visually, which of the two models is preferable under Occam's razor?
]
]

---

## R-squared

- $R^2$ is the percentage of variability in the response variable explained by 
the regression model.

```{r}
glance(book_main_fit)$r.squared
glance(book_int_fit)$r.squared
```

--
- Clearly the model with interactions has a higher $R^2$.

--
- However using $R^2$ for model selection in models with multiple explanatory variables is not a good idea as $R^2$ increases when **any** variable is added to the model.

---

## Adjusted R-squared

... a (more) objective measure for model selection

- Adjusted $R^2$ doesn't increase if the new variable does not provide any new 
informaton or is completely unrelated, as it applies a penalty for number of 
variables included in the model.
- This makes adjusted $R^2$ a preferable metric for model selection in multiple
regression models.

---

## Comparing models

.pull-left[
```{r}
glance(book_main_fit)$r.squared
glance(book_int_fit)$r.squared
```
]
.pull-right[
```{r}
glance(book_main_fit)$adj.r.squared
glance(book_int_fit)$adj.r.squared
```
]

--

.pull-left-wide[
.small[
- Is R-sq higher for int model?
```{r}
glance(book_int_fit)$r.squared > glance(book_main_fit)$r.squared 
```

- Is R-sq adj. higher for int model?

```{r}
glance(book_int_fit)$adj.r.squared > glance(book_main_fit)$adj.r.squared
```
]
]
---

class: middle

# More detail on multiple predictors

---
class: middle

# Two numerical predictors

---

## The data

```{r load-pp, message=FALSE}
pp <- read_csv(
  "data/paris-paintings.csv",
  na = c("n/a", "", "NA")
) %>%
  mutate(log_price = log(price))
```

---

## Multiple predictors

- Response variable: `log_price` 
- Explanatory variables: Width and height

```{r model-price-width-height}
pp_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_price ~ Width_in + Height_in, data = pp)
tidy(pp_fit)
```

---

##  Linear model with multiple predictors

```{r model-price-width-height-tidy, echo=FALSE}
tidy(pp_fit)
```

<br>

$$\widehat{log\_price} = 4.77 + 0.0269 \times width - 0.0133 \times height$$

---

## Visualizing models with multiple predictors

.panelset[
.panel[.panel-name[Plot]
.pull-left-wide[
```{r plotly-plot, echo = FALSE, fig.align="center", warning=FALSE}
p <- plot_ly(pp,
  x = ~Width_in, y = ~Height_in, z = ~log_price,
  marker = list(size = 3, color = "lightgray", alpha = 0.5, 
                line = list(color = "gray", width = 2))) %>%
  add_markers() %>%
  plotly::layout(scene = list(
    xaxis = list(title = "Width (in)"),
    yaxis = list(title = "Height (in)"),
    zaxis = list(title = "log_price")
  )) %>%
  config(displayModeBar = FALSE)
frameWidget(p)
```
]
]
.panel[.panel-name[Code]
```{r plotly-code, eval=FALSE}
p <- plot_ly(pp,
  x = ~Width_in, y = ~Height_in, z = ~log_price,
  marker = list(size = 3, color = "lightgray", alpha = 0.5, 
                line = list(color = "gray", width = 2))) %>%
  add_markers() %>%
  plotly::layout(scene = list(
    xaxis = list(title = "Width (in)"),
    yaxis = list(title = "Height (in)"),
    zaxis = list(title = "log_price")
  )) %>%
  config(displayModeBar = FALSE)
frameWidget(p)
```
]
]

---

class: middle

# Numerical and categorical predictors

---

## Price, surface area, and living artist

- Explore the relationship between price of paintings and surface area, conditioned 
on whether or not the artist is still living
- First visualize and explore, then model
- But first, prep the data

.midi[
```{r}
pp <- pp %>%
  mutate(artistliving = if_else(artistliving == 0, "Deceased", "Living"))

pp %>%
  count(artistliving)
```
]

---

## Typical surface area

.panelset[
.panel[.panel-name[Plot]
.pull-left-narrow[
Typical surface area appears to be less than 1000 square inches (~ 80cm x 80cm). There are very few paintings that have surface area above 5000 square inches.
]
.pull-right-wide[
```{r ref.label = "viz-surf-artistliving", echo = FALSE, warning = FALSE, out.width="90%"}
```
]
]
.panel[.panel-name[Code]
```{r viz-surf-artistliving, fig.show = "hide"}
ggplot(data = pp, aes(x = Surface, fill = artistliving)) +
  geom_histogram(binwidth = 500) + 
  facet_grid(artistliving ~ .) +
  scale_fill_manual(values = c("#E48957", "#071381")) +
  guides(fill = FALSE) +
  labs(x = "Surface area", y = NULL) +
  geom_vline(xintercept = 1000) +
  geom_vline(xintercept = 5000, linetype = "dashed", color = "gray")
```
]
]

---

## Narrowing the scope

.panelset[
.panel[.panel-name[Plot]
For simplicity let's focus on the paintings with `Surface < 5000`:

```{r ref.label = "viz-surf-lt-5000-artistliving", echo = FALSE, warning = FALSE, out.width = "55%"}
```
]
.panel[.panel-name[Code]
```{r viz-surf-lt-5000-artistliving, fig.show = "hide"}
pp_Surf_lt_5000 <- pp %>%
  filter(Surface < 5000)

ggplot(data = pp_Surf_lt_5000, 
       aes(y = log_price, x = Surface, color = artistliving, shape = artistliving)) +
  geom_point(alpha = 0.5) +
  labs(color = "Artist", shape = "Artist") +
  scale_color_manual(values = c("#E48957", "#071381"))
```
]
]

---

## Facet to get a better look

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "viz-surf-lt-5000-artistliving-facet", echo = FALSE, warning = FALSE, out.width = "80%", fig.asp = 0.5}
```
]
.panel[.panel-name[Code]
```{r viz-surf-lt-5000-artistliving-facet, fig.show = "hide"}
ggplot(data = pp_Surf_lt_5000, 
       aes(y = log_price, x = Surface, color = artistliving, shape = artistliving)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~artistliving) +
  scale_color_manual(values = c("#E48957", "#071381")) +
  labs(color = "Artist", shape = "Artist")
```
]
]

---

## Two ways to model

- **Main effects:** Assuming relationship between surface and logged price 
**does not vary** by whether or not the artist is living.
- **Interaction effects:** Assuming relationship between surface and logged 
price **varies** by whether or not the artist is living.

---

## Interacting explanatory variables

- Including an interaction effect in the model allows for different slopes, i.e. 
nonparallel lines.
- This implies that the regression coefficient for an explanatory variable would 
change as another explanatory variable changes.
- This can be accomplished by adding an interaction variable: the product of two 
explanatory variables.

---

## Two ways to model

.pull-left-narrow[
- **Main effects:** Assuming relationship between surface and logged price **does not vary** by whether or not the artist is living
- **Interaction effects:** Assuming relationship between surface and logged price **varies** by whether or not the artist is living
]
.pull-right-wide[
```{r pp-main-int-viz, echo=FALSE, out.width="70%", fig.asp = 0.9}
pp_main_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_price ~ Surface + artistliving, data = pp_Surf_lt_5000)
pp_main_fit_aug <- augment(pp_main_fit$fit)

pp_int_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_price ~ Surface * artistliving, data = pp_Surf_lt_5000)
pp_int_fit_aug <- augment(pp_int_fit$fit)

p_main <- ggplot(
  pp_main_fit_aug,
  aes(y = log_price, x = Surface, color = artistliving)
) +
  geom_point(aes(shape = artistliving), alpha = 0.5) +
  scale_x_continuous(breaks = c(0, 2500, 5000)) +
  scale_color_manual(values = c("#E48957", "#071381")) +
  geom_line(aes(y = .fitted), size = 1.5) +
  labs(y = "log_price", title = "Main effects", color = "Artist", shape = "Artist")

p_int <- ggplot(
  pp_int_fit_aug,
  aes(y = log_price, x = Surface, color = artistliving)
) +
  geom_point(aes(shape = artistliving), alpha = 0.5) +
  scale_x_continuous(breaks = c(0, 2500, 5000)) +
  scale_color_manual(values = c("#E48957", "#071381")) +
  geom_line(aes(y = .fitted), size = 1.5) +
  labs(y = "log_price", title = "Interaction effects", color = "Artist", shape = "Artist")

p_main /
  p_int  + 
  plot_layout(guides = "collect") & theme(legend.position = "bottom")
```
]

---

## Fit model with main effects

- Response variable: `log_price`
- Explanatory variables: `Surface` area and `artistliving`

.midi[
```{r model-main-effects}
pp_main_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_price ~ Surface + artistliving, data = pp_Surf_lt_5000)
tidy(pp_main_fit)
```
]

--

$$\widehat{log\_price} = 4.88 + 0.000265 \times surface + 0.137 \times artistliving$$

---

## Solving the model

- Non-living artist: Plug in 0 for `artistliving`

$\widehat{log\_price} = 4.88 + 0.000265 \times surface + 0.137 \times 0$  
$= 4.88 + 0.000265 \times surface$

--
- Living artist: Plug in 1 for `artistliving`

$\widehat{log\_price} = 4.88 + 0.000265 \times surface + 0.137 \times 1$   
$= 5.017 + 0.000265 \times surface$

---

## Visualizing main effects

.pull-left-narrow[
- **Same slope:** Rate of change in price as the surface area increases does 
not vary between paintings by living and non-living artists.
- **Different intercept:** Paintings by living artists are consistently more 
expensive than paintings by non-living artists.
]
.pull-right-wide[
```{r out.width="100%", echo = FALSE}
p_main
```
]

---

## Interpreting main effects

.midi[
```{r exp-coefs}
tidy(pp_main_fit) %>% 
  mutate(exp_estimate = exp(estimate)) %>%
  select(term, estimate, exp_estimate)
```
]

- All else held constant, for each additional square inch in painting's surface area, the price of the painting is predicted, on average, to be higher by a factor of 1.
- All else held constant, paintings by a living artist are predicted, on average, to be higher by a factor of 1.15 compared to paintings by an artist who is no longer alive.
- Paintings that are by an artist who is not alive and that have a surface area of 0 square inches are predicted, on average, to be 132 livres.

---

## Main vs. interaction effects

- The way we specified our main effects model only lets `artistliving` affect the intercept.
- Model implicitly assumes that paintings with living and deceased artists have the *same slope* and only allows for *different intercepts*.  

.question[
What seems more appropriate in this case?
+ Same slope and same intercept for both colours
+ Same slope and different intercept for both colours
+ Different slope and different intercept for both colours
]

---

## Interaction: `Surface * artistliving`

```{r out.width="60%", echo = FALSE}
p_int
```

---

## Fit model with interaction effects

- Response variable: log_price
- Explanatory variables: `Surface` area, `artistliving`, and their interaction

.midi[
```{r model-interaction-effects}
pp_int_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_price ~ Surface * artistliving, data = pp_Surf_lt_5000)
tidy(pp_int_fit)
```
]

---

## Linear model with interaction effects

.midi[
```{r model-interaction-effects-tidy, echo=FALSE}
tidy(pp_int_fit)
```
]

$$\widehat{log\_price} = 4.91 + 0.00021 \times surface - 0.126 \times artistliving$$
$$+ ~ 0.00048 \times surface * artistliving$$

---

## Interpretation of interaction effects

- Rate of change in price as the surface area of the painting increases does 
vary between paintings by living and non-living artists (different slopes), 
- Some paintings by living artists are more expensive than paintings by
non-living artists, and some are not (different intercept).

.small[
.pull-left[
- Non-living artist: 
$\widehat{log\_price} = 4.91 + 0.00021 \times surface$
$- 0.126 \times 0 + 0.00048 \times surface \times 0$
$= 4.91 + 0.00021 \times surface$
- Living artist: 
$\widehat{log\_price} = 4.91 + 0.00021 \times surface$
$- 0.126 \times 1 + 0.00048 \times surface \times 1$
$= 4.91 + 0.00021 \times surface$
$- 0.126 + 0.00048 \times surface$
$= 4.784 + 0.00069 \times surface$
]
.pull-right[
```{r viz-interaction-effects2, out.width="80%", echo = FALSE}
p_int
```
]
]

---

## Comparing models

It appears that adding the interaction actually increased adjusted $R^2$, so we 
should indeed use the model with the interactions.

```{r}
glance(pp_main_fit)$adj.r.squared
glance(pp_int_fit)$adj.r.squared
```

---

## Third order interactions

- Can you? Yes
- Should you? Probably not if you want to interpret these interactions in context
of the data.
