---
title: "HW 4 - Model and review"
subtitle: "Due: 18 November, 12:00 UK time"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Notes

------------------------------------------------------------------------

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)

knitr::opts_chunk$set(out.width = "100%", eval = TRUE)
```

# Part 1 - General Social Survey

## Scientific research

```{r read_data}
gss16<-read.csv("data/gss16.csv")
```

### Ex. 1. a)

```{r}
gss16 <- gss16 %>%
  mutate(
    advfront = case_when(
      advfront %in% c("Agree", "Strongly agree") ~ "Agree",
      is.na(advfront) ~ NA_character_,
      TRUE ~ "Not agree"
    ),
    advfront = fct_relevel(advfront, "Not agree", "Agree")
  )

gss16 %>%
  count(advfront)

```

### Ex. 1. b)

```{r}
gss16 <- gss16 %>%
  mutate(
    polviews = case_when(
      str_detect(polviews, "[Cc]onservative") ~ "Conservative",
      str_detect(polviews, "[Ll]iberal") ~ "Liberal",
      TRUE ~ polviews
    ),
    polviews = fct_relevel(polviews, "Conservative", "Moderate", "Liberal")
  )

gss16 %>%
  count(polviews)
```

### Ex. 1. c)

```{r}
gss16_advfront <- gss16 %>%
  select(advfront, educ, polviews, wrkstat) %>%
  drop_na()

gss16_advfront[1:10,]
```

### Ex. 2.

```{r}
set.seed(1234)
gss16_split <- initial_split(gss16_advfront)
gss16_train <- training(gss16_split)
gss16_test  <- testing(gss16_split)
```


#### recipe

```{r}
gss16_rec_1 <- recipe(advfront ~ polviews + wrkstat + educ, data = gss16_train) %>%
  step_other(wrkstat, threshold = 0.10, other = "Other") %>%
  step_dummy(all_nominal(), -all_outcomes())
```

#### model

```{r}
gss16_spec <- logistic_reg() %>%
  set_engine("glm")
```

#### workflow

```{r}
gss16_wflow_1 <- workflow() %>%
  add_model(gss16_spec) %>%
  add_recipe(gss16_rec_1)
```

#### Narrative 
A logistic regression is the appropriate model to use here because the response variable is binary (agree or not agree).


### Ex. 3.

```{r}
gss16_fit_1 <- gss16_wflow_1 %>%
  fit(gss16_train)
tidy(gss16_fit_1)

```

```{r}
o1<-exp(tidy(gss16_fit_1)$estimate[1]+13*tidy(gss16_fit_1)$estimate[2])

o2<-exp(tidy(gss16_fit_1)$estimate[1]+13*tidy(gss16_fit_1)$estimate[2]+tidy(gss16_fit_1)$estimate[4])

o1
o2

# the odds ratio
o2/o1

# the exponential of the coefficient associated with liberal political views

exp(tidy(gss16_fit_1)$estimate[4])
```


#### Narrative

$o_1$ are the predicted odds for someone who is conservative, has an education level of 13 and is retired agreeing with the statement. $o_2$ are the predicted odds for someone who is liberal, has an education level of 13 and is retired agreeing with the statement. Thus the ratio $o_2/o_1$ is the odds ratio. **The odds ratio tells us that all else being equal, the odds of someone with liberal political views agreeing with the statement are 1.97 times greater than the odds of someone with conservative views agreeing with the statement.** This is also how we can interpret the exponential of the coefficients. 


### Ex. 4. 

Define workflow for the second model and fit it
```{r message=FALSE}
gss16_rec_2 <- recipe(advfront ~ polviews + educ, data = gss16_train) %>%
  step_dummy(all_nominal(), -all_outcomes())

gss16_wflow_2 <- workflow() %>%
  add_model(gss16_spec) %>%
  add_recipe(gss16_rec_2)

gss16_fit_2 <- gss16_wflow_2 %>%
  fit(gss16_train)

```

Use both models to predict the test set, produce ROC curves and compute the area under the ROC curves.  

```{r fig.height=5, fig.width=5, warning=FALSE, out.width="50%", dpi = 300}


gss16_test_pred_1 <- predict(gss16_fit_1, new_data = gss16_test, type = "prob") %>%
  bind_cols(gss16_test %>% select(advfront))

gss16_test_pred_1 %>%
  roc_curve(truth = advfront, .pred_Agree, event_level = "second") %>%
  autoplot()

gss16_test_pred_1 %>%
  roc_auc(truth = advfront, .pred_Agree, event_level = "second")

gss16_test_pred_2 <- predict(gss16_fit_2, new_data = gss16_test, type = "prob") %>%
  bind_cols(gss16_test %>% select(advfront))

gss16_test_pred_2 %>%
  roc_curve(truth = advfront, .pred_Agree, event_level = "second") %>%
  autoplot()

gss16_test_pred_2 %>%
  roc_auc(truth = advfront, .pred_Agree, event_level = "second")

```

#### Narrative

Model 2 has better test set AUC (area under ROC curve) than model 1, although whether this performance is a substantive improvement is arguable. Never-the-less, as model 2 has fewer variables, which is generally favoured due to reduced model complexity, run time, and better generalisability, this model should be favoured.

