---
title: "HW 01 - All about Edinburgh"
author: "KEY"
date: "`r Sys.Date()`"
output: html_document
---

```{r load-packages, message = FALSE}

# Loading necessary data R packages
library(tidyverse)
library(lubridate)

# Loading necessary data sets from csv files
edibnb <- read_csv("data/edibnb.csv")
council <- read_csv("data/council_assessments.csv")
```

### Exercise 1

_Question_:

In the `edibnb` data set, 

- which variables have the missing observations, i.e. **NA**s as an entry? State the name of variables

**_Answer_:**

```{r hw01-ex1-01, message=F, warning=F}
# Option 1
edibnb %>% 
  select(where(~ any(is.na(.)))) %>% 
  summarise(across(everything(), ~sum(is.na(.))))

# Option 2
# This code will give you a summary of the number of NA values for each column that contains at least one NA value.
edibnb %>%
  summarise(across(where(~ any(is.na(.))), ~sum(is.na(.))))

```


```{r}
# Alternative option without counting the number of missing values 
# Focus on the variables having TRUE as output
edibnb %>% 
  summarise_all(anyNA)
```

In general, the variables *price*, *neighbourhood*, *bathrooms*, *bedrooms* , *beds* and *review_scores_rating* have different numbers of NAs in the data set. To illustrate, *neighborhood* variable includes *2294* missing observations whereas *bedrooms* variable has only *4* missing values.

**NOTE:** If students used simply the functions `sum()` and `is.na()` to detect **NA**s in a different way, still acceptable if they detected the variable names correctly. 

_Question_:

- what is the ID of the listing that has the highest number of reviews with a perfect review score of 100%?

**_Answer_:**

```{r hw01-ex1-02}
edibnb %>% 
  filter(review_scores_rating == 100) %>%
  arrange(desc(number_of_reviews)) %>%
  select(id, review_scores_rating, number_of_reviews) %>%
  slice(1)
```

The listing with a perfect review rating of 100% with the most number of reviews (612 reviews) is property ID 860170.

------------------------------------------------------------------------------

### Exercise 2

_Question_:

Find the number of Airbnb properties located in **Old Town** having the one night stay price is larger than 100 GBP.

**_Answer_:**

```{r}
edibnb %>%
  filter(
    neighbourhood == "Old Town", 
    price > 100,
  ) %>% count()
```

In total, 494 properties are detected based on the given condition. 

_Question_:

Calculate the minimum, maximum and average price from the Airbnb properties in Southside for a single night stay for four people. (Note, four people can stay in a property for four or more people.)

**_Answer_:**

```{r hw01-ex2}
edibnb %>%
  filter(
    accommodates >= 4,
    neighbourhood == "Southside"
  ) %>%
  summarise(
    min_price = min(price),
    max_price = max(price),
    av_price = mean(price)
  )
```

The data set is first filtered based on two conditions, that the property is in the Southside neighbourhood (`neighbourhood == "Southside"`) and that the listing accommodates at least 4 people (`accommodates >= 4`). The average price of listings that meet these conditions is £153 (to the nearest pound) with a min-max range of £10--£800.

------------------------------------------------------------------------------

### Exercise 3

_Question_:

Create a frequency table for the number of `bathrooms` in the data set for the properties located in **Newington**. 

**_Answer_:**

```{r}
edibnb %>% 
  group_by(neighbourhood) %>% 
  filter(neighbourhood == 'Newington') %>% 
  count(bathrooms)
  
```

Based on the created frequency table, most of the properties have 1 bathrooms (348 times). There is one missing observation (NA) in the table with value for the number of bathrooms that is not a whole number in some cases, e.g. 0.5 (3 times) 1.5 (48 times) or 3.5 (3 times). 

**NOTE:** The intention of using `group_by()` is about presenting the name of **Newington** in the output. However, without `group_by()` function, applying the `filter()` only and counting the number of observations is also suitable. If you used the second option directly, there is no deduction in your mark. 

_Question_:

Mutate the `bathrooms` variable to round the number of bathrooms _up_ to the nearest whole number (hint, look at `ceiling()`). Using this mutated variable, how many listings have more bathrooms than bedrooms?

**_Answer_:**

```{r hw01-ex3}
edibnb %>%
  mutate(whole_bathroom_number = ceiling(bathrooms)) %>%
  filter(whole_bathroom_number > bedrooms) %>%
  count()
```

The `ceiling()` function is used to round a number _up_ to the closest whole number. For example, `ceiling(4.0) == 4`, `ceiling(6.2) == 7` and `ceiling(9.9) == 10`. The above code uses the `ceiling()` function to mutate the `bathroom` variable so that the values for the number of bathrooms is a whole number, forming the new variable `whole_bathroom_number`. The data set is then filtered to extract the listing where the `whole_bathroom_number` is strictly greater than the number of `bedrooms`. The resulting data frame consists of 963 properties.

------------------------------------------------------------------------------

### Exercise 4

_Question_:

Using the **lubridate** package to correct the class of the `assessment_date` variable, and use this variable to evaluate how many properties the council assessed in the month of May.

**_Answer_:**

```{r hw01-ex41}
council %>%
  mutate(assessment_date = dmy(assessment_date)) %>%
  filter(month(assessment_date) == 5) %>%
  count()
```

The `dmy()` function from the **lubridate** package processes the `assessment_date` variable from a character data type to a recognisable date class. Furthermore, the `month()` function takes a date variable and extracts the month as a numerical variable (i.e. Jan = 1,  Feb = 2, ..., Dec = 12.) Filtering the `council` data set identifies that there were 296 assessments performed in the month of May.

_Question_:

Create a bar plot of the `neighbourhood` variable for the properties that have been assessed by the council (remember to iterate the data visualisation to make it as informative as possible). What does the data visualisation suggest? Is the council targeting all neighborhoods within Edinburgh equally?

**_Answer_:**

```{r hw01-ex42}
council %>% 
  left_join(edibnb, by="id") %>%
  ggplot(mapping = aes(y = neighbourhood))+
  geom_bar() + 
  labs(
    x = "Count",
    y = "Edinburgh Neighbourhood",
    title = "Number of AirBnB assessments per Edinburgh neighbourhood"
  )
```

It is clear to see that the council has perform many more assessments (1543 times) in Leith compared to the other neighbourhoods in Edinburgh.

**Note:** Other ways of joining the `edibnb` and `council` data sets that achieves the same result for this exercise are:

```{r alt_joins, eval = FALSE}
council %>% left_join(edibnb, by="id")
edibnb %>% right_join(council, by="id")
council %>% inner_join(edibnb, by = "id")
edibnb %>% inner_join(council, by = "id")
edibnb %>% semi_join(council, by = "id")
```
