---
title: "hw-01-key"
author: "Your Name (S000000)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cherryblossom)
```



## Exercise 1

#### Question
Look at the data dictionary. How many runners are in the data set? How many variables do we have for each runner? What type of variables are `net_sec` and `event`?


#### Answer

The `run17` data set contains details of **19,961** runners (number of rows) and has information across **9** variables for each runner.

The variable `net_sec` describes the time it takes for a runner to complete the race, after accounting for the staggered start, in seconds. Therefore the `net_sec` variable is a **continuous numerical (or quantitative) variable**.


The variable `event` indicates the race distance the runner participated in, whether that be the 10 mile race or the 5K (5,000 meters) race. This variable is a **categorical quantitative variable** because it is descriptive from a finite set of options.






## Exercise 2

#### Question

Amend the code to account for the following changes:
1.   Change the style of the data visualisation to a box-plot
2.   Edit the visualisation to make side-by-side box-plots (not above-below) based on the runner's identified Gender (`sex` variable).
3.   Facet the data visualisation depending on the runner's choose distance (`event` variable)
4.   Add an appropriate title and axis labels.

#### Answer

```{r run17_data_viz, warning = FALSE, out.width = "80%"}
ggplot(data = run17, 
       mapping = aes(x = sex,
                     y = net_sec)) +
  geom_boxplot() + 
  facet_wrap(~ event) + 
  labs(
    title = "Race times of the 2017 Cherry Blossom Run",
    x = "Identified Gender",
    y = "Race Time (in seconds)"
  )
```

#### Supplementary information not requested

Details about the changes made to the code for each amendment:

1.    Replaced the `geom_histogram()` function with `geom_boxplot()`.
2.    Changed the existing `net_sec` variable to the `y` and add `x = sex` into the aesthetics function. Note, keeping `x = net_sec` and adding `y = sex` will instead produce stacked (above - below) rather than side-by-side (left - right) box-plots.
3.    Add the `facet_wrap()` function with the `event` variable. An acceptable alternative command would be `facet_grid(~ event)`.
4.    Add the `labs()` function with the `title`, `x` and `y` inputs for specifying the title, x-axis label and y-axis label respectively. The text used for titles and labels should be brief yet descriptive about the data/variables.





## Exercise 3

#### Question

Give a brief description (about 2-3 sentences) about the structure of the data and the relationships between the variables in the data visualisation you have created in Exercise 2.

#### Answer

Some of the key features demonstrated by the data visualisation in Exercise 2 are:

-   The net time for runners who participated in the 5K race are broadly faster than those who did the 10 mile race.
-   Within each distance, the range of times is broadly similar between the male (M) and female (F) runners.
-   For both distances, the central estimate (mean/median) appears to be lower (i.e. faster race times) for the male runners.
-   The shape of the data for the 5K racerunners appears to be right (positively) skewed due to the number of outlier points at longer race times. On the contrary, the shape of the data appears to be symmetric for runners in the 10 Mile race.




## Exercise 4

#### Question

Find a _bad_ data visualisation and give at least _two_ reasons why there is a problem with the image.


#### Answer

Reasons for why a data visualisation is bad is potentially endless and will highly dependent on the specifics of the chosen image. Some reasons why a data visualisation is bad include:

-   There is a geometric flaw in the presentation of the data (e.g., bar heights not consistent with the data)
-   An overemphasis on making the visualisation 'pretty' at the sacrifice of the visualisation being informative
-   An ambiguity in how to understand the visualisation (e.g. two point types in the legend with the same appearance that describe different scenarios)
-   Overloading the visualisation to incorporate too many variables
-   Poor demonstration of basic mathematics (e.g. percentages not adding to 100%)
-   No title or axis label, or axis labels that are incomprehensible
-   Axes containing no values or present values on an inconsistent scale (e.g. values of 10, 20, 75 and 80 appearing at equal distances along the axis)
-   The visualisation has design elements that may inadvertently emphasise stereotypes. 









## Exercise 5

#### Question

Find a visualisation that you think provides a _good_ representation of the data and give at least _two_ reasons to justify why you believe the visualisation is good.

#### Answer

As with exercise 4, potential reasons for why a data visualisation is good will highly dependent on the specifics of the chosen image. Some reasons why a data visualisation is bad include:

-   The use of an appropriate visualisation style (box-plot, histogram, scatter plot, etc.) for the variable types.
-   Informative use of axis labels and title
-   Good use of additional visualisation techniques (colour, point shape, faceting) to add extra variables that does not overload the image.
-   The data visualisation is clean and the intended reader can immediately understand the structure of the data.
-   Visualisation contains design elements that are inclusive (e.g. colour blindness)







