---
title: "Lab 00 - Hello IDS!"
author: "KEY"
date: "`r Sys.Date()`"
output: html_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

### Exercises 16-17

We set up the data frame. Edit the data below to include information about your team members.

```{r setup-tibble}
team_data <- tibble(
  name = c("Peter Rabbit", "Bugs Bunny", "White Rabbit", 
           "Judy Hopps", "Thumper"),
  github_username = c("mcgregors-garden", "whats-up-doc", "Im-late", 
                      "have-to-try", "say-nothing"), 
  colour = c("powderblue", "darkorange", "white", 
             "seagreen1", "plum"),
  hobby = c("Hopping", "Comedy", "Tailoring", 
            "Driving", "Drumming"),
  favourite_number = c(2, 8, 5, 7, 10)
  )

team_data
```

### Exercise 18

We now work on visualising our data. 

```{r bar-plot}
p1 <- ggplot(data = team_data, 
             mapping = aes(x = github_username, 
                           y = favourite_number,
                           fill = colour), 
             ) +
   geom_col() +
   scale_fill_identity()
p1
```

The code generates a bar plot if each member's favourite number with the colour of each bar representing the team member's favourite colour.


### Exercises 19

Change the text of the plot title and axis labels.

```{r labelled-bar-plot}
p1 + labs(x = "GitHub Username", 
          y = "Favourite Number",
          title = "Data collected from team Rabbit")
```

The command `labs()` allows us to manually specify the axis labels and plot title for the data visualisation.


### Exercise 20

```{r new-bar-plot}
p2 <- ggplot(team_data, aes(x = hobby)) +
  geom_bar()
p2
```

This bar plot is not a good data visualisation because each person has a different hobby. If we collected more data from everybody on the course then there will be many more different hobbies (i.e. more values along the x-axis) with relatively few repeats (i.e. not much vertical variation).